---
title: "Data Manipulation AFR"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(openxlsx)
```

```{r parametrization}
# User defined Attributes
Pop.values = "ITA"

# Input paths
match.path = "E:\\Ace2\\"

# Outpaths
geno.path = "E:\\Ace2\\Elaborati\\SNP Genotipi\\"
```

```{r udf}

```

## Population Manipulation

Il primo chunk ricrea una dataframe con le sequenze di interesse. Molto lungo per cui lo collasso.

```{r SNP BASE}
# Build SNP
TSI = data.frame(rs35803318  = c(rep("C/C", 70), rep("C/T", 6), 
                                 rep("T/T", 0), rep("C",85), rep("T", 7)),
                 rs147311723 = c(rep("G/G", 76), rep("G/A", 0),
                                 rep("A/A", 0), rep("G",92), rep("A", 0)),
                 rs41303171  = c(rep("T/T", 75), rep("T/C", 1),
                                 rep("C/C", 0), rep("C",1), rep("T", 91)),  
                 rs149039346 = c(rep("A/A", 76), rep("G/A", 0),
                                 rep("G/G", 0), rep("A", 92), rep("G", 0)),
                 rs777410473 = c(rep("T/T", 76), rep("T/C", 0),
                                 rep("C/C", 0), rep("T", 92), rep("C", 0)),
                 rs200180615 = c(rep("C/C", 76), rep("C/T", 0),
                                 rep("T/T", 0), rep("C", 92), rep("T", 0)),
                 rs199951323 = c(rep("A/A", 76), rep("A/C", 0),
                                 rep("C/C", 0), rep("A", 92), rep("C", 0)),
                 rs183135788 = c(rep("T/T", 76), rep("T/C", 0),
                                 rep("C/C", 0), rep("T", 92), rep("C", 0)),
                 rs774469453 = c(rep("A/A", 75), rep("A/-", 0),
                                 rep("-/-", 1), rep("A", 92), rep("-", 0)),
                 rs756905974 = c(rep("T/T", 76), rep("T/C", 0),
                                 rep("C/C", 0), rep("T", 92), rep("C", 0)),
                 rs202137736 = c(rep("T/T", 76), rep("T/C", 0),
                                 rep("C/C", 0), rep("T", 92), rep("C", 0)),
                 rs779199005 = c(rep("A/A", 76), rep("A/G", 0),
                                 rep("G/G", 0), rep("A", 92), rep("G", 0)),
                 rs748701949 = c(rep("C/C", 76), rep("C/T", 0),
                                 rep("T/T", 0), rep("C", 92), rep("T", 0)),
                 rs140473595 = c(rep("C/C", 76), rep("C/T", 0),
                                 rep("T/T", 0), rep("C", 92), rep("T", 0)),
                 rs200973492 = c(rep("A/A", 76), rep("A/G", 0),
                                 rep("G/G", 0), rep("A", 92), rep("G", 0)),
                 rs191860450 = c(rep("T/T", 76), rep("T/C", 0),
                                 rep("C/C", 0), rep("T", 92), rep("C", 0)),
                 rs199569050 = c(rep("G/G", 76), rep("G/A", 0),
                                 rep("A/A", 0), rep("G", 92), rep("A", 0)),
                 rs773676270 = c(rep("T/T", 75), rep("T/C", 1),
                                 rep("C/C", 0), rep("T", 92), rep("C", 0)),
                 rs147464721 = c(rep("G/G", 76), rep("G/A", 0),
                                 rep("A/A", 0), rep("G", 92), rep("A", 0)),
                 rs138390800 = c(rep("T/T", 76), rep("T/C", 0),
                                 rep("C/C", 0), rep("T", 92), rep("C", 0)),
                 rs185525294 = c(rep("T/T", 76), rep("T/A", 0),
                                 rep("A/A", 0), rep("T", 92), rep("A", 0)),
                 rs780574871 = c(rep("C/C", 76), rep("C/T", 0),
                                 rep("T/T", 0), rep("C", 92), rep("T", 0)),
                 rs749750821 = c(rep("C/C", 76), rep("C/T", 0),
                                 rep("T/T", 0), rep("C", 92), rep("T", 0)),
                 rs745514718 = c(rep("C/C", 76), rep("C/T", 0),
                                 rep("T/T", 0), rep("C", 92), rep("T", 0)),
                 rs372272603 = c(rep("G/G", 76), rep("G/A", 0),
                                 rep("A/A", 0), rep("G", 92), rep("A", 0)),
                 rs779790336 = c(rep("C/C", 76), rep("C/G", 0),
                                 rep("G/G", 0), rep("C", 92), rep("G", 0)),
                 rs2285666   = c(rep("C/C", 57), rep("C/T", 15),
                                 rep("T/T", 4), rep("C", 80), rep("T", 12)),
                 rs528054982 = c(rep("A/A", 76), rep("A/T", 0),
                                 rep("T/T", 0), rep("A", 92), rep("T", 0)),
                 rs201900069 = c(rep("C/C", 76), rep("C/T", 0),
                                 rep("T/T", 0), rep("C", 92), rep("T", 0)),
                 rs199804629 = c(rep("C/C", 76), rep("C/T", 0),
                                 rep("T/T", 0), rep("C", 92), rep("T", 0)),
                 rs768558803 = c(rep("G/G", 76), rep("G/A", 0),
                                 rep("A/A", 0), rep("G", 92), rep("A", 0)),
                 rs761614932 = c(rep("G/G", 76), rep("G/A", 0),
                                 rep("A/A", 0), rep("G", 92), rep("A", 0)),
                 rs73635825  = c(rep("A/A", 76), rep("A/G", 0),
                                 rep("G/G", 0), rep("A", 92), rep("G", 0)),
                 rs765966061 = c(rep("G/G", 76), rep("G/T", 0),
                                 rep("T/T", 0), rep("G", 92), rep("T", 0))
                 )
```

```{r SNP other}
TSI = TSI %>%
  mutate(Sample     = paste(Pop.values, 1:168, sep = ""),
         Population = Pop.values)
```

Per ottenere il sesso mi conviene usare ifelse sul long format, che avrei dovuto comunque ottenere. Quindi lo creo.

```{r Long and Wide format}
# Columns needed (no sample, no population)
SNP = names(TSI[, -c(35, 36)])

# Created a long format
TSI.long = TSI %>%
  pivot_longer(cols = SNP, 
               names_to  = "Location",
               values_to = "Genotype")

# Gender column
TSI.long = TSI.long %>%
  mutate(Gender = ifelse(str_length(Genotype) == 1, "M", "F"))

# reorder
TSI.long = TSI.long[, c(1,2,5,3,4)]

# Back to wide
TSI.wide = TSI.long %>%
  pivot_wider(id_cols = NULL,
              names_from  = "Location",
              values_from = "Genotype")

# cleaning
remove(TSI, SNP)
```


```{r matcher manipulation}
# Matcher import
ref.al = read.xlsx(paste(match.path, "Reference alleles.xlsx", sep = ""), sheet = 3)

# Recode the var name from Variant to Location 
# It has to match with data.long
ref.al = ref.al %>%
  rename(Location = Variant)
```

## Allelic counts

```{r Allelic counts}
# Count 1 every time an allele occours: es. T = 1, T/C = 1/1, T/T = 2
# Del stands for Deletion, in this scenario deletion are considered like an allele
TSI.long = TSI.long %>%
  mutate(A = str_count(string = Genotype, pattern = "A"),
         C = str_count(string = Genotype, pattern = "C"),
         G = str_count(string = Genotype, pattern = "G"),
         "T" = str_count(string = Genotype, pattern = "T"),
         Del = str_count(string = Genotype, pattern = "-"))

# Merge data.long with Ref.al by Location so the Ref allele matchs with alleles
TSI.long = left_join(TSI.long, ref.al, by = "Location")

# Craft Al.1 column
TSI.long = TSI.long %>% 
  mutate(Al.1 = ifelse(REF == "A", A, 
                ifelse(REF == "C", C,
                ifelse(REF == "G", G,
                ifelse(REF == "T", `T`,
                ifelse(REF == "-", Del, NA))))))

# Craft Al.2 colum
TSI.long = TSI.long %>%
  mutate(Al.2 = ifelse(ALT == "A", A,
                ifelse(ALT == "C", C,
                ifelse(ALT == "G", G,
                ifelse(ALT == "T", `T`,
                ifelse(ALT == "-", Del, NA))))))

# check if there are some missing values due to errors
# You can do this check because original files have NOT NA
sum(is.na(TSI.long))
```


```{r Al conts object}
# Summarising data.long in a matrix of counts for downstream analysis
Al.counts = TSI.long %>%
  group_by(Location) %>%
  summarise(Al.1 = sum(Al.1),
            Al.2 = sum(Al.2))

# Recreate a Population column for sake of clarity
Al.counts = Al.counts %>%
  mutate(Population = paste(Pop.values))

# Reorder
Al.counts = Al.counts[, c(4,1:3)]
```

## Genotype Counts

Ricordati che se un maschio ha uno SNP sull'x e ha genotipo A e una femmina ha lo stesso SNP e ha genotipo A/A, se il ref. Allele è A allora entrambi sono omozigoti wild type. 

Quindi nel conteggio dei genotipi A del maschio e A/A della femmina valgono 1 entrambi.

```{r Genotype Counts}
# Removing vars used for Al.counts
Gen.counts = TSI.long[, -c(6:10, 13, 14)]

# Creating Homozygote Reference counts' column
# Given that there are SNP on X males have only 1 allele so we need to account that
# duplicating for the case when Gen = A and you're male and the case when Gen =A/A
Gen.counts = Gen.counts %>% 
  mutate(Homozygote_Reference = ifelse(REF == "A" & Genotype == "A", 1,
                                ifelse(REF == "C" & Genotype == "C", 1,
                                ifelse(REF == "G" & Genotype == "G", 1,
                                ifelse(REF == "T" & Genotype == "T", 1,
                                ifelse(REF == "A" & Genotype == "A/A", 1,
                                ifelse(REF == "C" & Genotype == "C/C", 1,
                                ifelse(REF == "G" & Genotype == "G/G", 1,
                                ifelse(REF == "T" & Genotype == "T/T", 1, 0)))))))))

# Homo Variant column
Gen.counts = Gen.counts %>% 
  mutate(Homozygote_Variant = ifelse(ALT == "A" & Genotype == "A", 1,
                              ifelse(ALT == "C" & Genotype == "C", 1,
                              ifelse(ALT == "G" & Genotype == "G", 1,
                              ifelse(ALT == "T" & Genotype == "T", 1,
                              ifelse(ALT == "A" & Genotype == "A/A", 1,
                              ifelse(ALT == "C" & Genotype == "C/C", 1,
                              ifelse(ALT == "G" & Genotype == "G/G", 1,
                              ifelse(ALT == "T" & Genotype == "T/T", 1, 0)))))))))

# Hetero column
Gen.counts = Gen.counts %>% 
  mutate(Heterozygote = ifelse(Homozygote_Reference == 0 & Homozygote_Variant == 0, 1, 0))

# Check for NA values // IF >0 the computation was wrong
sum(is.na(Gen.counts))
```

```{r Gen.count Object}
# Summarise to SNP level
Gen.counts = Gen.counts %>%
  group_by(Location) %>%
  summarise(Homozygote_Reference = sum(Homozygote_Reference),
            Heterozygote = sum(Heterozygote),
            Homozygote_Variant = sum(Homozygote_Variant))

# Create population column
Gen.counts = Gen.counts %>%
  mutate(Population = Pop.values)

# reordering
Gen.counts = Gen.counts[, c(5, 1:4)]
```

## Export for Genotype Data

In questo chunk esporto i genotipi, le conte alleliche e genotipiche e il dataset in versione long per completezza

```{r}
# Create export list
gen.list = list(Genotypes         = TSI.wide,
                "Allelic counts"  = Al.counts,
                "Genotype counts" = Gen.counts,
                "Check for Long"  = TSI.long)

# Export automat
write.xlsx(gen.list, file = paste(geno.path, paste(Pop.values, ".xlsx", sep = ""),sep = ""))
```